Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-12-01 presentations/illustrations/presentation-illustrations.tex
Added slides for linear capabilities and attack 2. Cleaned up some tikz code.

@@

\begin{document}
\begin{frame}
  \frametitle{Traditional stack pointers}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=white] (0,-.5) rectangle (6,2.5) node[pos=.5] {\footnotesize higher stack frames...};
    \end{scope}
    \draw[->] (-1,0) -- node[midway,sloped,above] {stack grows upward} (-1,15);

@@

\begin{frame}
  \frametitle{Stack and return capabilities}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \begin{scope}
        \clip (-.1,-.1) rectangle (6.1,15);
        \draw[fill=white] (0,2.5) rectangle (6,15.1);
      \end{scope}
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

\begin{frame}
  \frametitle{Stack and return capabilities: Attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \begin{scope}
        \clip (-.1,-.1) rectangle (6.1,15);
        \draw[fill=white] (0,2.5) rectangle (6,15.1);
      \end{scope}
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

    % stores stack pointer in heap
    \begin{onlyenv}<+->
      \begin{scope}
        \clip (13.9,6) rectangle (18.1,13);
        \draw[fill=gray!20] (14,5.9) rectangle (18,13.1);
      \end{scope}
      \draw (16,14) node {heap memory};
      \draw[fill=gray!50] (14,8) rectangle node[color=red] {\scriptsize copy of old sp} (18,8.5);
    \end{onlyenv}
    \draw<.> (14,8.25) edge[->,red,thick,bend left] (sp1);
    \begin{onlyenv}<+->
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (11.5,6.5) -- (11.5,16) node[draw=black] (spold) [black,midway,xshift=0.8cm] {};
      \draw (14,8.25) edge[->,red,thick,bend left] (spold);
    \end{onlyenv}

@@

\begin{frame}
  \frametitle{Stack and return capabilities: Attack 2}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \draw[fill=white] (0,4) rectangle (6,16);
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

\begin{frame}
  \frametitle{Local Stack Capabilities prevent Attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \draw[fill=white] (0,9.5) rectangle (6,15.2);
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

    % stores stack pointer in heap
    \begin{scope}
      \clip (13.9,6) rectangle (18.1,13);
      \draw[fill=gray!20] (14,5.9) rectangle (18,13.1);
    \end{scope}
    \draw (16,14) node {heap memory};
    \draw[fill=gray!50] (14,8) rectangle node[color=red] {\scriptsize copy of old sp} (18,8.5);
    \draw (14,8.25) edge[->,red,thick,bend left] (sp1);
    \draw[teal,very thick] (13.5,7.5) -- (18.5,9);
    \draw[teal,very thick] (18.5,7.5) -- (13.5,9);
    \draw node[teal,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw, fill=white] () at (14,4) {Stack pointer is local!};  
\end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Local caps: Stack clearing prevents Attack 2}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \draw[fill=white] (0,4) rectangle (6,16);
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

\begin{frame}
    \frametitle{Linear caps: stack pointers}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw (0,0) -- (0,15);
      \draw (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

\begin{frame}
    \frametitle{Linear caps prevent attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw (0,0) -- (0,15);
      \draw (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

@@

    \begin{onlyenv}<.->
      \begin{scope}
        \clip (13.9,9) rectangle (18.1,14);
        \draw[fill=gray!20] (14,8.9) rectangle (18,14.1);
      \end{scope}
    \end{onlyenv}
    \draw<.>[fill=white] (14,10) rectangle node[color=red] {} (18,10.5);

@@

    \draw<+>[teal,very thick] (6.4,4.4) -- (7.4,5.4);
    \draw<.>[teal,very thick] (7.4,4.4) -- (6.4,5.4);
    \draw<.> node[teal,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw, fill=white, align=center] () at (14,4) { Splice fails,\\ stack pointer is linear!};

>>>>

\newcommand{\stdstackstart}{
  \scope
    \clip (-.1,-.1) rectangle (6.1,15.1);
    \fill[fill=white] (0,0) rectangle (6,15);
    \draw (0,0) -- (0,15);
    \draw (6,0) -- (6,15);
    \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
  \endscope
  \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);
}

@@

\begin{document}
\begin{frame}
  \frametitle{Traditional stack pointers}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \fill[gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=white] (0,-.5) rectangle (6,2.5) node[pos=.5] {\footnotesize higher stack frames...};
    \end{scope}
    \draw[->] (-1,0) -- node[midway,sloped,above] {stack grows upward} (-1,15);

@@

\begin{frame}
  \frametitle{Stack and return capabilities}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

\begin{frame}
  \frametitle{Stack and return capabilities: Attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    % stores stack pointer in heap
    \begin{onlyenv}<+->
      \begin{scope}
        \fill[gray!20] (14,6) rectangle (18,13);
        \draw (14,6) -- (14,13);
        \draw (18,6) -- (18,13);
      \end{scope}
      \draw (16,14) node {heap memory};
      \draw[fill=gray!50] (14,8) rectangle node[color=red] {\scriptsize copy of old sp} (18,8.5);
    \end{onlyenv}
    \draw<.> (14,8.25) edge[->,red,thick,bend left] (sp1);
    \begin{onlyenv}<+->
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (11.5,6.5) -- (11.5,16) node[draw=black] (spold) [black,midway,xshift=0.8cm] {};
      \draw (14,8.25) edge[->,red,thick,bend left] (spold);
    \end{onlyenv}

@@

\begin{frame}
  \frametitle{Stack and return capabilities: Attack 2}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

\begin{frame}
  \frametitle{Local Stack Capabilities prevent Attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    % stores stack pointer in heap
    \begin{scope}
      \fill[gray!20] (14,6) rectangle (18,13);
      \draw (14,6) -- (14,13);
      \draw (18,6) -- (18,13);
    \end{scope}
    \draw (16,14) node {heap memory};
    \draw[fill=gray!50] (14,8) rectangle node[color=red] {\scriptsize copy of old sp} (18,8.5);
    \draw (14,8.25) edge[->,red,thick,bend left] (sp1);
    \draw[teal,very thick] (13.5,7.5) -- (18.5,9);
    \draw[teal,very thick] (18.5,7.5) -- (13.5,9);
    \draw node[teal,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw, fill=white] () at (14,4) {Stack pointer is local!};  
\end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Local caps: Stack clearing prevents Attack 2}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

\begin{frame}
    \frametitle{Linear caps: stack pointers}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

\begin{frame}
    \frametitle{Linear caps prevent attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

@@

    \begin{onlyenv}<.->
      \begin{scope}
        \fill[gray!20] (14,9) rectangle (18,14);
        \draw (14,9) -- (14,14);
        \draw (18,9) -- (18,14);
      \end{scope}
    \end{onlyenv}
    \draw<.>[fill=white] (14,10) rectangle node[color=red] {} (18,10.5);

@@

    \draw<+>[teal,very thick] (6.4,4.4) -- (7.4,5.4);
    \draw<.>[teal,very thick] (7.4,4.4) -- (6.4,5.4);
    \draw<.> node[teal,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw, fill=white, align=center] () at (14,4) { Splice fails,\\ stack pointer is linear!};  
  \end{tikzpicture}  
\end{frame}

\begin{frame}
  \frametitle{Linear caps prevents attack 2}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \stdstackstart

    % Callee called
    \begin{onlyenv}<+>
      \draw [decorate,decoration={brace,aspect=0.2,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6.5,5) -- (6.5,16) node[draw=black] (usp) [black,pos=0.2,xshift=0.8cm] {};
      \capbrace[sp2]{(6.5,2.5)}{(6.5,5)}
      \draw node[right=0cm of usp] { stack pointer };
      \draw node[fill=gray!50, right=0cm of sp2] { return data pointer };
      \inactsf{(0,2.5)}{(6,5)}{caller stack frame} 
      \actadv{(0,5)}{(6,8)}{callee stack frame}
    \end{onlyenv}

    \begin{onlyenv}<+-.(1)>
      \capbrace[sp2]{(6.5,2.5)}{(6.5,5)}
      \draw node[fill=gray!50, right=0cm of sp2] { return data pointer };

      \capbrace[usp]{(13.5,5)}{(13.5,16)}
      \draw (6,13.25) edge[red,thick,->,in=45] (usp);
      \draw node[right=0cm of usp] { stack pointer };

      \draw[fill=white] (0,13) rectangle (6,13.5) node[pos=.5,color=red] {\footnotesize sp};
      \inactsf{(0,2.5)}{(6,5)}{caller stack frame} 
      \actadv{(0,5)}{(6,8)}{callee stack frame}
    \end{onlyenv}

    \draw<.> node[teal,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw, fill=white, align=center] () at (4,10) { Stack pointer\\essentially lost.};

    \begin{onlyenv}<+->
      \capbrace[sp2]{(6.5,2.5)}{(6.5,5)}
      \draw node[right=0cm of sp2] { return data pointer };

      \capbrace[usp]{(13.5,5)}{(13.5,16)}
      \draw (6,13.25) edge[red,thick,->,in=45] (usp);
      \draw node[right=0cm of usp] { stack pointer };

      \begin{scope}
        \fill[fill=gray!50] (0,2.5) rectangle (6,15) node[pos=.5,color=black] {};
        \draw (0,0) -- (0,15);
        \draw (6,0) -- (6,15);
      \end{scope}
      \draw[fill=gray!50] (0,13) rectangle (6,13.5) node[pos=.5,color=red] {\footnotesize sp};
      \actsf{(0,2.5)}{(6,5)}{caller stack frame} 
  \end{onlyenv}

  \draw<+>[teal,very thick] (6.4,4.4) -- (7.4,5.4);
  \draw<.>[teal,very thick] (7.4,4.4) -- (6.4,5.4);
  \draw<.> node[teal,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw, fill=white, align=center] () at (14,7) { Splice fails, no\\stack pointer returned.};
